---
title: "RSDC: Regime-Switching Correlation Models in R"
date: "`r Sys.Date()`"
abstract: >
  The \CRANpkg{RSDC} package implements regime-switching correlation models
  in R, extending the @pelletier2006 framework with time-varying transition
  probabilities (TVTP) driven by exogenous covariates and with an arbitrary
  number of regimes through a multinomial-logit (softmax) parameterization.
  Estimation combines a global search in a reparameterized space in which
  every candidate is a valid correlation matrix with local refinement and a
  multi-start replication diagnostic, making the maximum-likelihood
  estimator usable in higher cross-sectional dimensions. We present the
  methodology, the package design, a Monte Carlo validation, and a strictly
  out-of-sample study of five Fama-French industry portfolios in which
  transitions respond to two nearly orthogonal covariates, the Media Climate
  Change Concerns index and the VIX, and feed regime-based minimum-variance
  and maximum-diversification allocations. All results are reproducible from
  data shipped with the package.
draft: false
author:
  - name: David Ardia
    affiliation: CIRANO, GERAD, HEC Montréal
    address:
      - Department of Decision Science
      - Montréal, Canada
    email: david.ardia@hec.ca
    orcid: 0000-0003-2823-782X
  - name: Benjamin Seguin
    affiliation: HEC Montréal
    address:
      - Department of Decision Science
      - Montréal, Canada
  - name: Roosevelt Ymele Nguemo
    affiliation: HEC Montréal
    address:
      - Department of Decision Science
      - Montréal, Canada
    orcid: 0009-0003-9399-6256
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
    citation_package: natbib
    keep_tex: true
bibliography: RSDC.bib
editor_options:
  markdown:
    wrap: 72
header-includes:
  - |
    \usepackage{graphicx}
    \usepackage{float}
    \floatplacement{table}{H}
    \floatplacement{figure}{H}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "center", fig.pos = "H", fig.width = 7, fig.height = 3.6, dpi = 110,
  message = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE)
options(digits = 4)
# rjtools' post-knit package check queries the CRAN task views over a
# hardcoded http URL; force the working https mirror so the render does not
# stall on networks that block plain http.
if (requireNamespace("rjtools", quietly = TRUE)) {
  try(assignInNamespace(
    ".cran",
    function(path = "") paste0("https://cran.r-project.org", path),
    ns = "rjtools"), silent = TRUE)
}
# Namespaces loaded here (uncached) so cached chunks keep their S3 methods
library(RSDC)
library(rugarch)
library(ggplot2)
thm <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(colour = "grey45", size = 10),
        axis.title.x = element_blank(),
        strip.text = element_text(face = "bold", hjust = 0))
```

# Introduction

Modeling time-varying correlations between asset returns is central to
portfolio construction and risk management. Smooth-transition approaches
such as the DCC model of @engle2002 let correlations evolve continuously.
An alternative view is that correlations spend long spells near
regime-specific levels and move abruptly when the market changes state.
Regime-switching dynamic correlation (RSDC) models [@pelletier2006]
formalize this view: the covariance matrix is decomposed into univariate
volatilities and a correlation matrix $R_{S_t}$ driven by a latent Markov
state $S_t$, so the correlation matrix is constant within a regime and
shifts discretely between regimes.

The \CRANpkg{RSDC} package implements this model class and extends it
in two directions. First, transition probabilities may be *time-varying*
(TVTP): they respond to any observable indicator supplied by the user, a
volatility index, a sentiment or news-attention series, a macroeconomic
factor, following the state-dependent Markov-switching tradition of
@filardo1994, @diebold1994 and @gray1996. Second, all specifications
support an arbitrary number of regimes $N$, with a logistic link for
$N = 2$ and a multinomial-logit (softmax) link for $N \ge 3$. In the
empirical illustration the covariates are the Media Climate Change Concerns
index [@ardia2023] and the VIX, but nothing in the package is specific to
climate applications.

Three model variants are provided: a constant-correlation baseline
(`"const"`), the fixed-transition Markov-switching model of @pelletier2006
(`"noX"`), and the covariate-driven TVTP model (`"tvtp"`). Estimation is by
maximum likelihood through the Hamilton filter [@hamilton1989]. A
practical obstacle in this model class is that the likelihood must be
maximized over correlation matrices: a box of pairwise correlations almost
never contains a positive-definite point once $K$ grows, which starves any
box-constrained global optimizer. Since version 1.7-0 the package removes
this obstacle by searching in the canonical partial-correlation
parameterization of @joe2006generating, in which *every* point of the
search box is a valid correlation matrix; the full estimation strategy is described in
the Methodology section.

Several CRAN packages address neighbouring problems;
Table \@ref(tab:related) summarizes the comparison. \CRANpkg{rmgarch}
[@ghalanos2024rmgarch] implements the DCC family, in which correlations
evolve smoothly and transitions cannot be tied to covariates.
\CRANpkg{MSGARCH} [@ardia2019] provides Markov-switching GARCH for
univariate volatility. General hidden-Markov toolkits such as
\CRANpkg{depmixS4} [@visser2010] and \CRANpkg{MSwM} can represent
switching in generic parameters but offer neither correlation-matrix
regimes with guaranteed positive definiteness nor portfolio tooling.
\CRANpkg{RSDC} fills this gap: regime-switching of the correlation
matrix itself, covariate-driven transitions, and a direct pipeline from
correlation forecasts to minimum-variance and maximum-diversification
portfolios.

\small

```{r related, echo = FALSE, cache = FALSE}
rel <- data.frame(
  Package = c("RSDC", "rmgarch", "MSGARCH", "depmixS4", "MSwM"),
  `Model class` = c("RS correlation",
                    "DCC (smooth)",
                    "MS-GARCH",
                    "hidden Markov",
                    "MS regression"),
  `Corr. regimes` = c("yes", "no", "no", "generic", "no"),
  `Covar. transitions` = c("yes", "no", "no", "no", "no"),
  `Portfolio tools` = c("yes", "no", "no", "no", "no"),
  check.names = FALSE)
knitr::kable(rel, caption = "Related CRAN packages. `depmixS4` can represent
  switching in generic distribution parameters but has no notion of a
  positive-definite correlation-matrix regime.")
```

\normalsize

Throughout, the input data are *standardized residuals*: returns filtered
by univariate volatility models (e.g., GARCH) and scaled to unit variance,
so that the regime dynamics identified by the package pertain to
correlations only. The remainder of the paper presents, in turn, the methodology, the
package design, a simulation-based validation, and the empirical
illustration.

# Methodology {#sec:methodology}

## Regime-switching correlation model

Let $U_t$ denote the $K$-dimensional vector of standardized residuals at
time $t$ and $S_t \in \{1, \dots, N\}$ a latent regime indicator.
Conditional on the regime, $U_t$ is multivariate normal with a
regime-specific correlation matrix:

$$
U_t \mid (S_t = j) \sim \mathcal{N}(0, R_j),
$$

where each $R_j$ is a $K \times K$ correlation matrix with $K(K-1)/2$ free
parameters $\rho^{(j)}$. With $N = 1$ the model reduces to the
constant-correlation baseline (`"const"`). For identifiability, the
estimators order regimes by ascending mean correlation, so regime 1 is
always the lowest-correlation state; this labeling convention is used in
all outputs.

## Markov switching dynamics

The regime follows a first-order Markov chain. We write $P$ for a constant
transition matrix and $\pi_{ij}(t)$ for the operative transition
probability at time $t$: the constant entry $P_{ij}$ in the
fixed-transition model, or the covariate-dependent value defined below in
the TVTP model.

### Fixed transitions (`"noX"`)

The transition matrix $P$, with
$P_{ij} = \Pr(S_t = j \mid S_{t-1} = i)$, is constant. For $N = 2$,

$$
P =
\begin{pmatrix}
p_{11} & 1 - p_{11} \\
1 - p_{22} & p_{22}
\end{pmatrix},
$$

with $p_{11}, p_{22}$ the regime persistences. For $N \ge 3$, row $i$ is
parameterized by its first $N - 1$ probabilities, the last entry following
by complement; each free probability is bounded in $[0.01, 0.99]$.

### Time-varying transitions (`"tvtp"`)

With covariates $x_t \in \mathbb{R}^{p}$ (including an intercept column if
desired, the package does not add one automatically), the transition
matrix varies over time. For $N = 2$ the *stay* probabilities follow a
logistic link,

$$
\pi_{ii}(t) = \frac{\exp(x_t^\top \beta_{i})}{1 + \exp(x_t^\top \beta_{i})},
\qquad i = 1, 2,
$$

with the switching probability obtained by complement,
$\pi_{ij}(t) = 1 - \pi_{ii}(t)$ for $j \neq i$. Here $\beta_i$ is a
length-$p$ coefficient vector; when $x_t$ contains an intercept and a
single indicator, a large positive slope in $\beta_1$ means that regime 1
becomes more persistent when the indicator is high.

For $N \ge 3$ a multinomial-logit (softmax) link is used: for each current
state $i$,

$$
\pi_{ij}(t) =
\frac{\exp(x_t^\top \beta_{ij})}{\sum_{k=1}^{N} \exp(x_t^\top \beta_{ik})},
\qquad j = 1, \dots, N,
$$

with the $N$-th logit fixed at zero as the reference category, leaving
$N - 1$ free coefficient vectors $\beta_{ij}$ per row. The $N = 2$ logistic
link is this softmax specialized to two states with state $i$ itself as the
modeled category, so the two displays use one convention. The coefficient
index carries the difference: one vector $\beta_i$ per row when $N = 2$,
$N - 1$ vectors $\beta_{ij}$ per row when $N \ge 3$.

### Model variants and parameter count

The three variants differ only in their transition block. The
constant-correlation model (`method = "const"`) is the $N = 1$ special
case: a single matrix $R_1$ estimated over the whole sample, whose
log-likelihood admits a closed form; it is the natural baseline. The
fixed-transition model (`method = "noX"`), following @pelletier2006,
captures regime shifts without exogenous covariates. The TVTP model
(`method = "tvtp"`) lets regime durations depend explicitly on observed
covariates such as financial-stress indicators, volatility indices, or
news-attention series.

The number of estimated parameters is:

* **TVTP:** transition coefficients, $N \times p$ for $N = 2$ (logistic) or
  $N(N-1) \times p$ for $N \ge 3$ (softmax), plus $N$ correlation matrices,
  i.e. $N\,K(K-1)/2$ correlations in total;
* **Fixed transitions (`noX`):** the $N(N-1)$ free entries of $P$ plus the
  $N\,K(K-1)/2$ correlations;
* **Constant correlation (`const`):** a single correlation matrix,
  $K(K-1)/2$ parameters;

where $p$ is the number of columns of $x_t$, including the intercept if one
is supplied. Means and variances are not estimated: inputs are assumed
standardized, and the fitted objects report placeholder regime means (0)
and volatilities (1) that do not enter the likelihood. Model selection can
be guided by information criteria such as the AIC and BIC.

## Likelihood inference via the Hamilton filter

Parameters are estimated by maximum likelihood, with the likelihood
evaluated by the Hamilton filter [@hamilton1989]. The filter proceeds as
follows.

**Prediction step.** Given the filtered probability from the previous time
step, $\xi_{i,t-1} = \Pr(S_{t-1} = i \mid \Omega_{t-1})$, where
$\Omega_{t-1}$ is the information available up to time $t-1$, the
predictive probability of being in state $j$ at time $t$ is

$$
\Pr(S_t = j \mid \Omega_{t-1}) = \sum_{i=1}^{N} \xi_{i,t-1}\,\pi_{ij}(t),
$$

with $\pi_{ij}(t)$ the operative transition probability defined above.

**Update step.** We observe $U_t$ and compute its likelihood under each
regime,

$$
f(U_t \mid S_t = j) = (2\pi)^{-K/2}\, |R_j|^{-1/2}
\exp\!\left\{ -\tfrac{1}{2}\, U_t^\top R_j^{-1} U_t \right\},
$$

and Bayes' rule updates the filtered probabilities:

$$
\xi_{j,t} = \Pr(S_t = j \mid \Omega_t) =
\frac{\Pr(S_t = j \mid \Omega_{t-1})\, f(U_t \mid S_t = j)}
     {\sum_{m=1}^{N} \Pr(S_t = m \mid \Omega_{t-1})\, f(U_t \mid S_t = m)}.
$$

**Likelihood.** The log-likelihood accumulates the prediction-error
decomposition,

$$
\sum_{t=1}^{T} \log f(U_t \mid \Omega_{t-1}), \qquad
f(U_t \mid \Omega_{t-1}) =
\sum_{j=1}^{N} \Pr(S_t = j \mid \Omega_{t-1})\, f(U_t \mid S_t = j).
$$

**Initialization and smoothing.** At $t = 1$ the filter uses a uniform
prior $\xi_{i,0} = 1/N$; an optional warm-start state (`xi_init`) is used
internally by the two-pass out-of-sample forecast. After filtering, a
backward recursion delivers the smoothed probabilities
$\Pr(S_t = j \mid \Omega_T)$. The filter's inner loop is implemented in
C++ (via \CRANpkg{RcppArmadillo}), with the pure-R implementation retained
as a tested reference.

## A feasible-by-construction global search {#sec:estimation}

The likelihood surface of a switching model is multimodal, and its natural
parameter space is hostile to global search: a correlation matrix must be
positive definite, which is a *joint* constraint on the $K(K-1)/2$
pairwise correlations that no per-coordinate box can express. The
probability that a uniform draw from the box is a valid matrix collapses
with $K$, near zero already for $K \ge 7$, so a naive global
optimizer starts from an almost entirely infeasible population. Since
version 1.7-0 the package therefore searches in a reparameterized space in
which feasibility holds *by construction*, with three ingredients.

**Partial correlations.** The global stage explores the canonical partial
correlations of a C-vine [@joe2006generating]: a vector
$z \in (-1, 1)^{K(K-1)/2}$ whose map to correlation matrices is a
*bijection* with the set of positive-definite correlation matrices, every point of the box is a valid model at any $K$. The first $K - 1$
coordinates are plain correlations with the first variable, deeper
coordinates are partial correlations, so the parameterization remains
interpretable; it is the same transform used by Stan for its correlation
matrix type.

**Bounded-softmax transition head.** For `"noX"` with $N \ge 3$ the free
transition probabilities of each row face a second joint constraint (a row
must sum to at most one). The global stage searches per-row softmax logits
bounded at $\pm 4$ instead: every point of the box now yields a valid row,
and the bound, the historical $[0.01, 0.99]$ probability box mapped to
the log-odds scale, keeps every transition probability away from zero,
so no regime can degenerate into an unreachable state during the search.
The $N = 2$ head, which faces no joint constraint, is unchanged.

**Search, refinement, and replication.** Each search runs differential
evolution [@mullen2011; @ardia2011] in the reparameterized space, refines
the best mutually distant candidates of the final population by L-BFGS-B in
the natural space, and returns the best refined point. Because a single
stochastic search can still land in a suboptimal basin of a multimodal
surface, `control = list(n_starts = k)` runs $k$ independent searches from
derived seeds and keeps the best fit; the per-start log-likelihoods are
stored on the fit, and a warning is issued when the maximum is not
replicated by at least two searches. Data-driven warm starts
(`rsdc_starts()`) remain available as a deterministic complement: when both
routes agree, the optimum is cross-certified. All searches are
reproducible: the seed is exposed via `control$seed` (default 123) and
results are invariant to the number of cores.

Model selection across fitted specifications proceeds by information
criteria; because the three variants are nested (set covariate slopes to
zero to obtain `"noX"`; collapse regimes to obtain `"const"`),
likelihood-ratio tests are available with base R alone, as the empirical
illustration shows. A free by-product of nesting is a
consistency check: a nested model can never attain a higher maximized
likelihood than the model that contains it, so any violation reveals an
under-optimized fit.

# Package design and usage {#sec:design}

The user-facing API is small. Table \@ref(tab:api) lists the exported
functions; everything else (S3 methods, tidiers) follows R conventions.

\small

```{r api, echo = FALSE, cache = FALSE}
api <- data.frame(
  Function = c("rsdc_estimate()", "rsdc_starts()", "rsdc_likelihood()",
               "rsdc_hamilton()", "rsdc_forecast()", "rsdc_forecast_ahead()",
               "rsdc_corr_bands()", "rsdc_viterbi()", "rsdc_bootstrap()",
               "rsdc_simulate()", "rsdc_minvar(), rsdc_maxdiv()"),
  Purpose = c("fit const, noX or tvtp by maximum likelihood",
              "data-driven warm starts (multi-start)",
              "likelihood evaluation",
              "Hamilton filter and smoother",
              "correlation and covariance paths",
              "multi-step-ahead forecasts",
              "uncertainty bands on the correlation path",
              "most likely (MAP) regime path",
              "parametric bootstrap (SEs, CIs)",
              "simulate from a fitted or specified model",
              "long-only MV and MD portfolio weights"))
knitr::kable(api, caption = "Exported functions of RSDC.")
```

\normalsize

`rsdc_estimate()` takes `method`, `residuals`, `N`, `X`, `out_of_sample`
and a `control` list, and returns an `"rsdc_fit"` object storing, among others, `correlations` (the $N \times
K(K-1)/2$ regime correlations), `transition_matrix` (for `"tvtp"`,
evaluated at the in-sample covariate means, a point summary, not the
path), `beta`, `log_likelihood` (and `log_likelihood_oos` when
`out_of_sample = TRUE`), and filtered/smoothed probabilities. Standard
generics apply: `print`, `summary`, `coef`, `logLik`, `AIC`, `BIC`,
`nobs`, `vcov`, `confint`, `predict`, `simulate`, `plot`, plus
`broom`-style `tidy()`/`glance()`/`augment()` and a `ggplot2::autoplot()`
method. The `control` list exposes the estimation protocol: `seed`
(default 123), `n_starts`, `cores`, `itermax`/`NP`/`steptol` (global-stage
budget), `maxit` (refinement), `compute_se`, and `start` (a warm start,
which skips the global search).

As a compact walkthrough, and a first look at the data machinery used in
the empirical illustration, we fit the two-regime TVTP model to the
green/brown portfolio pair shipped with the package [@ardia2023], with the
Media Climate Change Concerns (MCCC) index as the transition covariate.
Returns are first filtered by univariate GJR-GARCH(1,1) models
(\CRANpkg{rugarch}); to keep the walkthrough strictly out-of-sample we
estimate every model on the first 70% of the sample
(`out_of_sample = TRUE`) and reserve the remainder for evaluation, scaling
the covariate by its training-window moments.

```{r gb-data}
library(RSDC)
library(rugarch)

data("greenbrown"); data("mccc")
stopifnot(identical(greenbrown$DATE, mccc$DATE))
gb_ret <- as.matrix(greenbrown[, c("return_green", "return_brown")]) * 100
n_gb   <- nrow(gb_ret)
cut_gb <- round(0.7 * n_gb)                    # the package's 70/30 split

# GJR-GARCH(1,1), zero mean, fitted on the training window only; the fitted
# parameters then filter the full series (no test-window information enters).
garch_filter <- function(ret, cut) {
  spec <- ugarchspec(
    variance.model = list(model = "gjrGARCH", garchOrder = c(1, 1)),
    mean.model     = list(armaOrder = c(0, 0), include.mean = FALSE))
  K <- ncol(ret)
  z <- s <- matrix(NA_real_, nrow(ret), K,
                   dimnames = list(NULL, colnames(ret)))
  for (i in seq_len(K)) {
    fit <- ugarchfit(spec, ret[, i], solver = "hybrid",
                     out.sample = nrow(ret) - cut)
    spec_i <- spec; setfixed(spec_i) <- as.list(coef(fit))
    filt <- ugarchfilter(spec_i, ret[, i])
    z[, i] <- as.numeric(residuals(filt)) / as.numeric(sigma(filt))
    s[, i] <- as.numeric(sigma(filt))
  }
  for (i in seq_len(K))                       # train-window scale
    z[, i] <- z[, i] / sd(z[seq_len(cut), i])
  list(z = z, s = s)
}
gb <- garch_filter(gb_ret, cut_gb)
mccc_tr  <- mccc$mccc
mccc_std <- (mccc_tr - mean(mccc_tr[1:cut_gb])) /
  sd(mccc_tr[1:cut_gb])
X_gb <- cbind(intercept = 1, MCCC = mccc_std)
```

```{r gb-fit}
fit_gb <- rsdc_estimate("tvtp", residuals = gb$z, N = 2, X = X_gb,
                        out_of_sample = TRUE,
                        control = list(n_starts = 4, cores = 4))
fit_gb
```

The printed fit reports the regime correlations (ordered low to high), the
transition summary, and both the training log-likelihood and the out-of-sample
predictive log-likelihood. `summary(fit_gb)` adds standard errors from the
observed information, delta-method standard errors for the implied
transition probabilities, and regime diagnostics (expected durations,
ergodic distribution); `confint()` provides Wald intervals. Regimes beyond
three are supported for `"noX"` and `"tvtp"` (`N = 4` and higher), with the
same softmax head.

# Validation {#sec:validation}

Two kinds of evidence support the implementation. First, an analytic
anchor: with `method = "const"` the maximum-likelihood correlation matrix
is the sample correlation of the standardized residuals, and the package
reproduces it to numerical precision (this identity is part of the test
suite). Second, a full Monte Carlo study ships with the package: the
script `inst/simulation/monte_carlo_study.R` regenerates it end to end,
and its report `montecarlo_validation.pdf` is installed alongside and
available online at
\url{https://github.com/ArdiaD/RSDC/blob/master/inst/simulation/montecarlo_validation.pdf}.
The study covers five specifications, constant,
`noX` and `tvtp` at $N \in \{2, 3\}$, simulated and re-estimated with
$M = 100$ replications per cell, $T \in \{500, 1000, 2000\}$ and, for the
three-regime fixed-transition case, $K \in \{2, 3, 5\}$.

Three findings matter for users. Bias and RMSE shrink with $T$ in every
cell, at the $\sqrt{T}$ rate where the surface is regular, and every cell
converges in all 100 replications, including $K = 5$ (33 parameters),
a dimension at which the pre-1.7-0 natural-space search produced no usable
replication at all. Median RMSE *falls* as $K$ grows (each added series is
one more observation on the same latent regime path), which turns the
conjecture that regime-switching correlation models pay off increasingly
with dimension into a measured result. Finally, in the three-regime TVTP
cells a fraction of replications reach the $\pm 10$ coefficient bound at
small $T$, the classical separation problem of multinomial-logit
transitions when some transitions are observed only a handful of times, and that fraction falls from 75% to 36% of replications as $T$ doubles;
the study's usable counts report it transparently.

# Empirical illustration {#sec:empirical}

The main application uses the `ff5ind` data set shipped with the package:
daily value-weighted returns (percent) of five Fama-French industry
portfolios, Manufacturing, Energy, High Tech, Health, Utilities
[@fama1997industry], together with the MCCC index (monthly Aggregate,
forward-filled across trading days) and the CBOE VIX, aligned on
$T = 5155$ U.S. trading days from 2005-01-03 to 2025-06-30. The two
candidate transition covariates are nearly orthogonal in this window,
which lets the data separate two mechanisms: market fear (VIX) and climate
attention (MCCC).

Everything below is out-of-sample in the strict sense: the GARCH
pre-filter, the covariate scaling, and all regime-model parameters are
estimated on the first 70% of the sample ($T_{\text{train}} = 3608$ days,
2005-2019); the remaining 30% ($T_{\text{test}} = 1547$ days, 2019-2025,
spanning the COVID crash and the 2022 tightening cycle) is used only for
evaluation, with out-of-sample correlations weighted by *filtered* (not
smoothed) probabilities so no future information enters.

## Data and preprocessing

```{r ff5-data}
data("ff5ind")
industries <- c("Manuf", "Enrgy", "HiTec", "Hlth", "Utils")
ret   <- as.matrix(ff5ind[, industries])
dates <- ff5ind$DATE
n     <- nrow(ret)
cut   <- round(0.7 * n)                        # matches the estimators' split
c(train_end = format(dates[cut]), test_start = format(dates[cut + 1]))

ff5 <- garch_filter(ret, cut)                  # train-window GARCH parameters
z <- ff5$z; s <- ff5$s

std_train <- function(x) (x - mean(x[1:cut])) / sd(x[1:cut])
X_mccc <- cbind(intercept = 1, MCCC   = std_train(ff5ind$mccc))
X_vix  <- cbind(intercept = 1, logVIX = std_train(log(ff5ind$vix)))
X_both <- cbind(X_mccc, logVIX = X_vix[, "logVIX"])
round(cor(X_mccc[, "MCCC"], X_vix[, "logVIX"]), 3)   # near-orthogonal covariates
```

## Estimation and model selection

We estimate nine specifications with the same call, all with
`out_of_sample = TRUE` and the multi-start protocol; per-model wall-clock
times are collected for the timing table below.

```{r ff5-fits}
ctrl <- list(n_starts = 4, cores = 14, compute_se = FALSE)
fit1 <- function(method, N, X = NULL) {
  t0 <- Sys.time()
  f  <- rsdc_estimate(method, residuals = z, N = N, X = X,
                      out_of_sample = TRUE, control = ctrl)
  attr(f, "minutes") <- as.numeric(Sys.time() - t0, units = "mins")
  f
}
fits <- list(
  "const"      = fit1("const", 1),
  "noX(N=2)"   = fit1("noX", 2),
  "noX(N=3)"   = fit1("noX", 3),
  "tvtp2-MCCC" = fit1("tvtp", 2, X_mccc),
  "tvtp2-VIX"  = fit1("tvtp", 2, X_vix),
  "tvtp2-both" = fit1("tvtp", 2, X_both),
  "tvtp3-MCCC" = fit1("tvtp", 3, X_mccc),
  "tvtp3-VIX"  = fit1("tvtp", 3, X_vix),
  "tvtp3-both" = fit1("tvtp", 3, X_both))
```

```{r ff5-comparison, cache = FALSE}
comparison <- data.frame(
  model    = names(fits),
  logLik   = sapply(fits, function(f) as.numeric(logLik(f))),
  npar     = sapply(fits, function(f) f$npar),
  AIC      = sapply(fits, AIC),
  BIC      = sapply(fits, BIC),
  logLikOOS = sapply(fits, function(f) f$log_likelihood_oos),
  row.names = NULL)
knitr::kable(comparison, digits = 1, caption = sprintf(
  "Model comparison, five Fama-French industries. Training window
   2005-%s (T = %d): log-likelihood, parameters, AIC, BIC. Last column:
   predictive log-likelihood on the out-of-sample window (T = %d), computed with
   training-window parameters and filtered probabilities.",
  format(dates[cut], "%Y"), cut, n - cut))
```

Three regimes dominate two on every criterion, and every switching model
dominates the constant baseline. The criteria disagree, however, on the
transition specification: the AIC selects the richest model (three regimes,
both covariates), the BIC, penalizing harder, prefers the
VIX-only variant, and the out-of-sample predictive log-likelihood is best for
the fixed-transition `noX(N=3)` model, with `tvtp3-both` a close second
(a 14 log-point gap over 1547 days). In-sample, the covariates add
substantial information about transitions; that edge does not translate
into better prediction on this particular hold-out, which opens with the
COVID crash, a regime entry unlike anything in the training window. We
retain `tvtp3-both`, the AIC choice, and the specification that lets us
study the covariate effects, as the working model, and report the
fixed-transition alternative alongside it in the portfolio evaluation.
Because the single-covariate models are nested in the two-covariate model,
likelihood-ratio tests need only base R:

```{r ff5-lr, cache = FALSE}
ll <- setNames(comparison$logLik, comparison$model)
np <- setNames(comparison$npar,   comparison$model)
lr <- function(big, small) {
  stat <- 2 * (ll[[big]] - ll[[small]]); df <- np[[big]] - np[[small]]
  c(LR = round(stat, 1), df = df,
    p = signif(pchisq(stat, df, lower.tail = FALSE), 2))
}
rbind("VIX adds to MCCC (N = 3)"  = lr("tvtp3-both", "tvtp3-MCCC"),
      "MCCC adds to VIX (N = 3)"  = lr("tvtp3-both", "tvtp3-VIX"),
      "three regimes vs two"      = lr("tvtp3-both", "tvtp2-both"))
```

The two covariates contribute asymmetrically on the training window: the
VIX adds overwhelming information given MCCC (LR = 100.7), while MCCC adds
significant but modest information given the VIX (LR = 14.6, p = 0.02).
Their near-orthogonality still matters, whatever MCCC contributes is not
market fear in disguise, and the bootstrap below shows that several
individual MCCC slopes are precisely estimated. (The three-vs-two-regime statistic is
indicative only, as regime-count tests face unidentified parameters under
the null; AIC, BIC and the out-of-sample likelihood all point the same way.) As
a consistency check, every nested pair in Table \@ref(tab:ff5-comparison)
satisfies the required likelihood inequality.

## Inference on the retained model

For inference we re-estimate the retained specification on the training
window (a warm start from the fit above, so no global search is repeated)
and run the parametric bootstrap.

```{r ff5-boot}
fit_ret <- rsdc_estimate("tvtp", residuals = z[1:cut, ], N = 3,
                         X = X_both[1:cut, ],
                         control = list(start = fits[["tvtp3-both"]]$par,
                                        compute_se = FALSE))
set.seed(123)
bs <- rsdc_bootstrap(fit_ret, B = 199, X = X_both[1:cut, ], seed = 123,
                     cores = 14)
```

The fitted object also carries observed-information standard errors for
every coefficient, delta-method standard errors for the implied transition
matrix at the covariate means, and regime diagnostics (expected durations,
ergodic distribution); the full `summary()` output is reported in the
Appendix.

```{r ff5-se, results = "hide"}
fit_se <- rsdc_estimate("tvtp", residuals = z[1:cut, ], N = 3,
                        X = X_both[1:cut, ],
                        control = list(start = fit_ret$par))
summary(fit_se)
```

The two inference routes agree: the observed-information standard errors
track the bootstrap ones closely (median ratio 1.09), and the two methods
classify 11 of the 12 covariate slopes identically at the 5 percent level,
the exception being the borderline calm-persistence MCCC slope. We keep the
bootstrap intervals as the headline, since several transition probabilities
sit near their bounds, where the quadratic approximation behind the
observed information is least reliable.

```{r ff5-boot-table, cache = FALSE}
ci <- bs$ci
rows <- c(grep("x2]", rownames(ci), fixed = TRUE),
          grep("x3]", rownames(ci), fixed = TRUE))
tab <- cbind(ci[rows, , drop = FALSE], se = bs$se[rows])
rownames(tab) <- sub("x2]", "MCCC]", sub("x3]", "logVIX]", rownames(tab)))
star <- ifelse(tab[, 1] * tab[, 2] > 0, "*", "")
knitr::kable(cbind(as.data.frame(round(tab, 3)), ` ` = star), caption = sprintf(
  "Percentile 95 percent bootstrap intervals for the MCCC and log-VIX transition
   slopes of the retained model (B = %d parametric replicates on the
   training window; coefficients are softmax logits with the crisis regime
   as reference; * = interval excludes zero).", bs$B))
```

Figures \@ref(fig:ff5-pers-vix) and \@ref(fig:ff5-pers-mccc) turn the
estimated coefficients into pictures; each curve varies one covariate with
the other held at its training mean. Against log-VIX, fear empties the
calm state and locks in the high-correlation ones: calm persistence falls
from about 0.7 in quiet markets to essentially zero at high VIX, while the
crisis regime's stay probability climbs from below 0.5 to near certainty.
Under stress the system does not merely visit the high-correlation states,
it stays there. Against MCCC, attention works mostly through the calm
state, whose persistence roughly quadruples across the observed range, so
durable low-correlation spells require high climate attention; crisis
persistence also rises somewhat. The two covariates thus act on different
margins of the transition matrix, consistent with each carrying its own
information in the likelihood-ratio tests. We read all of these as
conditional associations estimated from one sample, not causal effects.

```{r ff5-pers-fun}
# Stay probability pi_ii(x) from the softmax rows of the estimated beta
# (row i: two destination blocks of (intercept, MCCC, logVIX); ref = crisis)
stay_prob <- function(beta, i, x_mccc, x_vix) {
  x <- c(1, x_mccc, x_vix)
  l1 <- sum(x * beta[i, 1:3]); l2 <- sum(x * beta[i, 4:6])
  e <- exp(c(l1, l2, 0) - max(l1, l2, 0))
  (e / sum(e))[min(i, 3)]
}
persistence_fig <- function(vary = c("vix", "mccc")) {
  vary <- match.arg(vary)
  obs <- if (vary == "vix") X_vix[1:cut, 2] else X_mccc[1:cut, 2]
  grid <- seq(quantile(obs, 0.005), quantile(obs, 0.995), length.out = 200)
  B <- fit_ret$beta
  d <- do.call(rbind, lapply(1:3, function(i) data.frame(
    x = grid, regime = c("Calm", "Normal", "Crisis")[i],
    p = vapply(grid, function(g)
      if (vary == "vix") stay_prob(B, i, 0, g) else stay_prob(B, i, g, 0),
      0))))
  d$regime <- factor(d$regime, levels = c("Calm", "Normal", "Crisis"))
  lab <- d[ave(d$x, d$regime, FUN = max) == d$x, ]
  vj <- c(Calm = 0.5, Normal = 1.6, Crisis = -0.7)
  lab$vjust <- vj[as.character(lab$regime)]
  xlab_txt <- if (vary == "vix") "log VIX (standardized)" else
                                 "MCCC (standardized)"
  ggplot(d, aes(x, p, colour = regime)) +
    geom_line(linewidth = 0.9) +
    geom_text(data = lab, aes(label = regime, vjust = vjust),
              hjust = -0.15, fontface = "bold", size = 3.4,
              show.legend = FALSE) +
    geom_rug(data = data.frame(x = obs), aes(x = x), inherit.aes = FALSE,
             sides = "b", alpha = 0.15, length = unit(0.03, "npc")) +
    scale_colour_manual(values = c("#2166ac", "#e8a33d", "#d1541e"),
                        guide = "none") +
    scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
    labs(y = "P(stay in regime)", x = xlab_txt) +
    thm + theme(axis.title.x = element_text())
}
```

```{r ff5-pers-vix, out.width = "100%", fig.height = 3.9, fig.cap = "Regime persistence as a function of log-VIX, MCCC held at its training mean; TVTP(N = 3), training-window estimates. Ticks along the x-axis: observed log-VIX distribution."}
persistence_fig("vix")
```

```{r ff5-pers-mccc, out.width = "100%", fig.height = 3.9, fig.cap = "Regime persistence as a function of MCCC, log-VIX held at its training mean; TVTP(N = 3), training-window estimates. Ticks along the x-axis: observed MCCC distribution."}
persistence_fig("mccc")
```

Parameter uncertainty can also be propagated into the correlation path
itself. Figure \@ref(fig:ff5-band-fig) shows the 95 percent band produced
by `rsdc_corr_bands()`: parameter vectors are drawn from the estimated
sampling distribution of the training-window fit, the filter is rerun for
each draw over the full sample, and pointwise quantiles are formed; the
out-of-sample segment is shown for one economically interpretable pair,
old-economy Energy against new-economy HiTec. The band is tight inside
persistent regimes and widens sharply around regime switches, where
classification uncertainty dominates parameter uncertainty.

```{r ff5-bands}
bands <- rsdc_corr_bands(fit_se, residuals = z, X = X_both,
                         B = 200, seed = 123)
```

```{r ff5-band-fig, out.width = "100%", fig.height = 3.4, fig.cap = "Predicted Energy-HiTec correlation with a 95 percent parameter-uncertainty band (B = 200 draws from the training-window sampling distribution), out-of-sample segment."}
pair_names <- apply(utils::combn(industries, 2), 2, paste, collapse = "-")
b <- bands[[which(pair_names == "Enrgy-HiTec")]]
keep_oos <- (cut + 1):n
db <- data.frame(date = dates[keep_oos], fit = b[keep_oos, "fit"],
                 lo = b[keep_oos, "lower"], hi = b[keep_oos, "upper"])
ggplot(db, aes(date, fit)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#1f5fa8", alpha = 0.20) +
  geom_line(colour = "#1f5fa8", linewidth = 0.35) +
  labs(y = "Correlation (Enrgy-HiTec)") + thm
```

## Regimes and the out-of-sample correlation path

```{r ff5-forecast}
fc <- rsdc_forecast("tvtp", N = 3, residuals = z, X = X_both,
                    final_params = fits[["tvtp3-both"]], sigma_matrix = s,
                    value_cols = industries, out_of_sample = TRUE)
oos_dates <- dates[(cut + 1):n]
mean_corr_oos <- rowMeans(fc$predicted_correlations)
```

```{r ff5-smoothfig, echo = FALSE, out.width = "100%", fig.height = 5.4, fig.cap = "Smoothed regime probabilities of the retained model on the training window (top three panels; regimes ordered by mean correlation) and the mean predicted correlation on the out-of-sample window (bottom; filtered weights, 21-day moving average). The dashed line marks the train/test split."}
sm <- fit_ret$smoothed_probs
regime_means <- rowMeans(fit_ret$correlations)
labs <- sprintf("P(%s (rho~%.2f))", c("Calm", "Normal", "Crisis"), regime_means)
long <- data.frame(
  date = rep(dates[1:cut], 3),
  p    = as.vector(t(sm)),
  regime = factor(rep(labs, each = cut), levels = labs))
g_top <- ggplot(long, aes(date, p, colour = regime)) +
  geom_line(linewidth = 0.25) +
  facet_wrap(~regime, ncol = 1) +
  scale_colour_manual(values = c("#2166ac", "#e8a33d", "#d1541e"), guide = "none") +
  scale_y_continuous(breaks = c(0, 0.5, 1)) + labs(y = "Probability") + thm
d_oos <- data.frame(
  date = oos_dates,
  ma = as.numeric(stats::filter(mean_corr_oos, rep(1/21, 21), sides = 2)))
g_bot <- ggplot(d_oos, aes(date, ma)) +
  geom_line(colour = "#1f5fa8", linewidth = 0.5, na.rm = TRUE) +
  geom_vline(xintercept = as.numeric(dates[cut]), linetype = 2, colour = "grey55") +
  labs(y = "Mean corr. (OOS)") + thm
gridExtra::grid.arrange(g_top, g_bot, ncol = 1, heights = c(2.4, 1))
```

On the training window the three regimes read as calm, normal and crisis
states (mean correlations `r paste(round(regime_means, 2), collapse = " / ")`),
with the 2008-2009 and 2010-2012 episodes classified as crisis spells.
Out of sample, the filtered mean-correlation path rises sharply in March
2020 and again through 2022, periods the model has never seen, before easing over 2023-2025.

## Portfolio evaluation on the out-of-sample window

The out-of-sample covariances $\Sigma_t = D_t R_t D_t$ (training-window
GARCH parameters around filtered correlations) feed the long-only
minimum-variance and maximum-diversification optimizers, evaluated with
`lag = TRUE`: weights formed at $t-1$ earn the period-$t$ return, and
since parameters, covariate scaling and volatility filters all come from
the training window, the backtest is look-ahead-free in the strict sense.

```{r ff5-portfolio}
ret_oos <- ret[(cut + 1):n, ]
lagged_ret <- function(W) rowSums(rbind(NA, W[-nrow(W), ]) * ret_oos)
pf_pair <- function(fcst) {
  mv <- rsdc_minvar(sigma_matrix = fcst$sigma_matrix, value_cols = industries,
                    predicted_corr = fcst$predicted_correlations, y = ret_oos,
                    long_only = TRUE, lag = TRUE)
  md <- rsdc_maxdiv(sigma_matrix = fcst$sigma_matrix, value_cols = industries,
                    predicted_corr = fcst$predicted_correlations, y = ret_oos,
                    long_only = TRUE, lag = TRUE)
  list(mv = lagged_ret(mv$weights), md = lagged_ret(md$weights))
}
p_tvtp <- pf_pair(fc)
fc_noX <- rsdc_forecast("noX", N = 3, residuals = z,
                        final_params = fits[["noX(N=3)"]], sigma_matrix = s,
                        value_cols = industries, out_of_sample = TRUE)
p_noX <- pf_pair(fc_noX)
pf <- data.frame(`minvar (tvtp3-both)` = p_tvtp$mv,
                 `minvar (noX3)`       = p_noX$mv,
                 `maxdiv (tvtp3-both)` = p_tvtp$md,
                 equal = rowMeans(ret_oos), check.names = FALSE)[-1, ]
```

```{r ff5-pf-table, cache = FALSE}
pf_stats <- t(sapply(pf, function(r) {
  vol <- sd(r) * sqrt(252); ret_a <- mean(r) * 252
  cum <- cumprod(1 + r / 100)
  mdd <- -100 * max(1 - cum / cummax(cum))
  c(`vol (%)` = vol, `return (%)` = ret_a,
    `Sharpe` = ret_a / vol, `max DD (%)` = mdd)
}))
knitr::kable(round(pf_stats, 3), caption = sprintf(
  "Out-of-sample portfolio evaluation (%s to %s, %d days): annualized volatility,
   return and Sharpe ratio of the long-only strategies, minimum variance
   under the retained TVTP model and under the fixed-transition noX(N=3)
   alternative, maximum diversification under the retained model, against
   the equal-weight benchmark. Weights at t-1 earn the t return; all
   parameters are training-window estimates.",
  format(oos_dates[2]), format(max(oos_dates)), nrow(pf)))
```

```{r ff5-pf-fig, echo = FALSE, out.width = "100%", fig.height = 3.6, fig.cap = "Cumulative value of the strategies on the out-of-sample window (log scale)."}
cum <- data.frame(
  date = rep(oos_dates[-1], ncol(pf)),
  value = as.vector(apply(pf, 2, function(r) cumprod(1 + r / 100))),
  strategy = factor(rep(colnames(pf), each = nrow(pf)), levels = colnames(pf)))
ggplot(cum, aes(date, value, colour = strategy)) +
  geom_line(linewidth = 0.5) + scale_y_log10() +
  scale_colour_manual(values = c("#1f5fa8", "#7bb3e0", "#e8a33d", "grey45"),
                      name = NULL) +
  labs(y = "Cumulative value (log scale)") + thm +
  theme(legend.position = "bottom")
```

Both regime-based minimum-variance portfolios deliver a markedly lower
realized volatility than the equal-weight benchmark on data none of their
inputs have seen, at $K = 5$, where the full correlation matrix drives
the weights, and the two variants are nearly indistinguishable, so the
portfolio gain comes from the regime structure itself rather than from the
covariates. The volatility reduction is bought with lower average returns:
Sharpe ratios end up close to the benchmark's, slightly below it for the
minimum-variance portfolios. Minimum variance achieves exactly its stated
objective, no more.

## Timing

```{r ff5-timing, cache = FALSE}
timing <- data.frame(
  model   = names(fits),
  npar    = sapply(fits, function(f) f$npar),
  minutes = round(sapply(fits, function(f) attr(f, "minutes")), 1),
  row.names = NULL)
knitr::kable(timing, caption = sprintf(
  "Wall-clock estimation time per specification (four independent
   searches per fit on %d cores; T = %d training observations, K = %d). The C++
   filter evaluates one likelihood in about 2 ms at this sample size.",
  ctrl$cores, cut, ncol(z)))
```

Estimation cost is driven by the number of parameters through the global
stage's population size; the multi-start protocol runs its searches in
parallel, so its wall-clock cost is close to that of a single search.

# Conclusion

The \CRANpkg{RSDC} package estimates regime-switching correlation
models with constant, fixed-transition and covariate-driven transition
dynamics, for any number of regimes and, since version 1.7-0, in cross-
sectional dimensions where box-constrained search was previously
infeasible, the global stage operates in a partial-correlation space
where every candidate is a valid model [@joe2006generating]. A shipped
Monte Carlo study documents parameter recovery across all specifications
and shows accuracy *improving* with the cross-section; a five-industry
out-of-sample application favors three regimes on every criterion; within
the three-regime class, in-sample criteria and likelihood-ratio tests
support transitions driven by two nearly orthogonal indicators, market
fear associated with entries into the high-correlation regime, climate
attention predominantly with staying out of it, while the out-of-sample
predictive likelihood ranks the fixed-transition variant marginally ahead,
a reminder that in-sample transition fit need not convert into forecast
gains. The regime-based minimum-variance portfolios attain the lowest
out-of-sample volatility among the strategies considered, with or without
covariates.

Some limitations should be kept in mind. Regime labels rest on the
ascending-correlation ordering; all results are conditional on the
univariate GARCH pre-filter; TVTP coefficients can reach their bounds in
small samples when some transitions are rarely observed (the separation
issue quantified in the validation study); and the empirical exercise is a
software illustration on one asset universe, not a climate-finance study.
Within those bounds, the package provides a reproducible pipeline from
standardized residuals to regime-aware covariance forecasts and portfolio
weights, driven by data shipped with the package.

# Appendix: summary of the retained model {-}

The full `summary()` of the retained TVTP($N = 3$) model, estimated on the
training window, reports the softmax coefficients with observed-information
standard errors, the implied transition matrix at the covariate means with
delta-method standard errors, and per-regime diagnostics (stay
probability, expected duration, ergodic probability). The coefficient
naming follows `b[g<from>,c<softmax column>,x<covariate>]`, with $x_1$ the
intercept, $x_2$ the MCCC and $x_3$ the log-VIX.

\footnotesize

```{r ff5-se-appendix, echo = FALSE}
summary(fit_se)
```

\normalsize

# Computational details {-}

Results were obtained with \CRANpkg{RSDC} version
`r as.character(utils::packageVersion("RSDC"))`, using
\CRANpkg{rugarch} for GJR-GARCH filtering, \CRANpkg{DEoptim} for
the global stage, and \CRANpkg{quadprog} / \CRANpkg{Rsolnp} for
the portfolio optimizers. All computations use the default estimation seed
(123) and are reproducible from the packaged `greenbrown`, `mccc` and
`ff5ind` data sets.

# Acknowledgments {-}

We gratefully acknowledge financial support from the Fonds de recherche
AMF-GIRIF and the Natural Sciences and Engineering Research Council of
Canada (grant RGPIN-2022-03767).

