---
title: "Getting started with RSDC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RSDC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.4,
                      message = FALSE, warning = FALSE)
options(digits = 4)
```

```{r setup}
library(RSDC)
```

This vignette walks through the package API: estimating the three model
variants, using the standard S3 methods on a fitted object, comparing models with
`AIC`/`BIC`, forecasting, and simulating. To keep the build fast we use a small
simulated data set and a reduced optimiser budget via `control`; in real use, keep
the defaults.

## A fitted object is an `"rsdc_fit"`

```{r data}
set.seed(1)
n  <- 300L
X  <- cbind(intercept = 1, z = as.numeric(scale(seq_len(n))))   # include an intercept
beta <- rbind(c(1.5, -0.6),     # regime 1 (logistic stay-prob coefficients)
              c(1.0,  0.2))     # regime 2
rho  <- rbind(0.2, 0.8)         # K = 2 -> one correlation per regime
Sig  <- array(c(matrix(c(1, 0.2, 0.2, 1), 2),
                matrix(c(1, 0.8, 0.8, 1), 2)), dim = c(2, 2, 2))
sim  <- rsdc_simulate(n = n, X = X, beta = beta, mu = matrix(0, 2, 2),
                      sigma = Sig, N = 2, seed = 42)
y <- scale(sim$observations)    # treated as unit-variance standardized residuals

ctrl <- list(itermax = 30, NP = 40, seed = 1)   # small budget for the vignette
fit <- rsdc_estimate("tvtp", residuals = y, N = 2, X = X, control = ctrl)
class(fit)
fit
```

## Standard methods

`AIC()` / `BIC()` work directly because `logLik()` carries `df` and `nobs`:

```{r methods}
logLik(fit)
c(AIC = AIC(fit), BIC = BIC(fit), nobs = nobs(fit))
coef(fit)
```

`summary()` reports a coefficient table with standard errors (from the observed
information at the MLE), and `confint()` gives Wald intervals:

```{r summary}
summary(fit)
head(confint(fit))
```

## Model comparison

```{r compare}
fit_const <- rsdc_estimate("const", residuals = y, control = ctrl)
fit_noX   <- rsdc_estimate("noX",   residuals = y, N = 2, control = ctrl)
data.frame(
  model = c("const", "noX(N=2)", "tvtp(N=2)"),
  AIC   = c(AIC(fit_const), AIC(fit_noX), AIC(fit)),
  BIC   = c(BIC(fit_const), BIC(fit_noX), BIC(fit))
)
```

## Forecasting and simulation

`predict()` wraps `rsdc_forecast()`; supply the conditional volatilities:

```{r predict}
sigma_mat <- matrix(1, n, 2, dimnames = list(NULL, c("a", "b")))   # unit vols here
fc <- predict(fit, residuals = y, sigma_matrix = sigma_mat,
              value_cols = c("a", "b"), X = X)
str(fc$predicted_correlations)
```

`simulate()` draws a new path from the fitted model (here, TVTP given `X`):

```{r simulate}
sim2 <- simulate(fit, X = X, seed = 7)
table(sim2$states)
```

## More than three regimes

`N >= 4` is supported for `"noX"` and `"tvtp"`:

```{r bign}
fit4 <- rsdc_estimate("noX", residuals = y, N = 4,
                      control = list(itermax = 20, NP = 60, seed = 1))
dim(fit4$transition_matrix)
```
